---
title: "brGDGTNPP: Complete Workflow Tutorial"
author: "Vidal-Cordasco, M. and Manica, A."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brGDGTNPP: Complete Workflow Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Setup

```{r, eval = FALSE}
# Clear environment
rm(list = ls())

# Set working directory to where the script and package file are located
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()

# Load devtools for package development
library(devtools)
document("R_package")
build("R_package")

# Install and load the package
install.packages("brGDGTNPP_0.1.0.tar.gz", 
                 repos = NULL, 
                 type = "source")
library(brGDGTNPP)
library(magrittr)
```

# Load example data

```{r}
library(brGDGTNPP)
data(example_data)
head(example_data)
```

# Reliability/Applicability assessment

Check if new brGDGT data falls within the training data space.

```{r, warning = FALSE}
check_data <- diagnose_applicability(
  brgdt_data = example_data,
  plot = TRUE,
  threshold = 0.95,
  map = TRUE,
  n_neighbors = 5
)

# Results
print(check_data)
```

## Diagnostic plots

### PCA plot

```{r, fig.cap = "PCA of samples vs training data"}
check_data$plots$pca
```

### Mahalanobis distance plot

```{r, fig.cap = "Mahalanobis distances (statistical rarity)"}
check_data$plots$mahalanobis
```

### Geographic similarity map

```{r, fig.cap = "Training sites with most similar brGDGT composition"}
check_data$plots$similarity_map
```

### Ternary plot

```{r, fig.cap = "Methylation groups (training data sized by NPP)"}
check_data$plots$ternary
```

# Compute NPP

```{r}
results <- predict_npp(example_data, sample_type = "Lacustrine Sediment")
head(results)

# Save outputs
write.csv(results, "results.csv")
```

## Plot NPP with chronology

```{r, fig.cap = "NPP through time"}
# If your data include chronologies
results$Age <- example_data$Age

with(results, plot(x = Age, y = NPP_pred_mean, 
                   xlab = "Age BP", 
                   ylab = "Net Primary Productivity",
                   main = "Net Primary Productivity",
                   pch = 20,        
                   col = "black"))
```

# Potential NPP (Miami model)

Create climate data and compute potential NPP using the Miami model.

```{r}
# Create climate data with 50 samples
set.seed(123)

climate_data <- data.frame(
  ID = 1:50,
  MAT = seq(10, 15, length.out = 50) + rnorm(50, 0, 1),
  MAP = seq(400, 700, length.out = 50) + rnorm(50, 0, 100)
)

potential_npp <- miami_model(climate_data)
head(potential_npp)

# Plot results
with(potential_npp, plot(x = ID, y = NPP_pred_mean, 
                   xlab = "ID", 
                   ylab = "Net Primary Productivity",
                   main = "Potential NPP (Miami model)",
                   pch = 20,        
                   col = "blue"))
```

# Compute Human Appropriation of NPP (HANPP)

```{r}
hanpp <- miami_model(climate_data,
                    npp_observed = results$NPP_pred_mean[1:50],
                    hanpp = TRUE)

with(hanpp, plot(x = ID, y = HANPP, 
                         xlab = "ID", 
                         ylab = "HANPP",
                         main = "Human Appropriation of NPP",
                         pch = 20,        
                         col = "black"))
```

## Combined plot

```{r, fig.cap = "Comparison of NPP estimates"}
# Order ID values
hanpp_sorted <- hanpp[order(hanpp$ID), ]
potential_sorted <- potential_npp[order(potential_npp$ID), ]
results_sorted <- results[order(results$ID), ]

# Set y limits
y_limits <- range(c(hanpp_sorted$HANPP, 
                    potential_sorted$NPP_pred_mean, 
                    results_sorted$NPP_pred_mean), na.rm = TRUE)

# Create plot
plot(x = hanpp_sorted$ID, y = hanpp_sorted$HANPP,
     xlab = "ID", 
     ylab = "NPP (g/m²/yr)",
     main = "Net Primary Productivity",
     pch = 20,        
     col = "black",
     ylim = y_limits,
     type = "b")  

# Climate-derived NPP
lines(x = potential_sorted$ID, y = potential_sorted$NPP_pred_mean, 
      col = "blue", type = "b", pch = 20)

# brGDGT-derived NPP
lines(x = results_sorted$ID, y = results_sorted$NPP_pred_mean, 
      col = "red", type = "b", pch = 20)

# Legend
legend("topleft",
       legend = c("Climate-derived NPP", "brGDGT-derived NPP", "Difference (HANPP)"),
       col = c("blue", "red", "black"),
       pch = 20,
       lty = 1)
```

# Session information

```{r}
sessionInfo()
```
